Separating Use and Reuse to Improve Both

### 2017-05-04 FACS/Intro.tex
Intro

Historically, there has been a lot of focus on the importance of
separing subtype from subclassing~\cite{cook}.  This is claimed to be
good for code reuse, design and reasoning.  While there is no problem
in subtyping without subclassing, in most OO languages code reuse
(inheritance/extends) implies subtyping in a fundamental way. 
Consider the following (Java) code:

@@

Class \lstinline{B} inherits the method \lstinline{ma} from
\lstinline{A}.  We can see an invocation of \lstinline{ma} inside
\lstinline{mb}, where \lstinline{this} is of type \lstinline{B}.  The
execution will eventually call \lstinline{Util.m} with a \lstinline{B}
instance. This can be correct only if \lstinline{B} is a subtype of
\lstinline{A}. Thus, every OO language with the minimal features
exposed in this example (using \lstinline{this}, extends and method calls) is forced
to accept that subclassing implies subtyping.\footnote{ Note how this
  is true also in C++, where is possible to "extends privately". Such
  is a limitation of the visibility of subtyping but not over
  subtyping itself, and the former example would be accepted by C++
  even if \lstinline{B} was to "privately extends" \lstinline{A}.}
\marco{Cite some work of bruceTHisType and show how he also fails.}
\bruno{I was expecting some text/argument here on why subclassing
  implies subtyping is bad. Perhaps another example where subclassing
  must not imply subtyping to type-check.}

An iconic example on why connecting inheritance/code reuse and
subtpying is problematic is provided by the
historic\cite{LaLonde:1991:SSS:110673.110679}: A reasonable
implementation for a \Q@Set@ may be easy to extend into a \Q@Bag@
keeping tracks of how many times an element occurs.  We would just add
some state and override a couple of methods.\bruno{Are we going to
  present this example in the paper solved in 42, for example? 
I think I would expect to see it.}

However, our subclassing would break Liskov substitution principle (LSP)~\cite{martin2000design}: not all bags are sets!
Of course, one could retroactivelly fix this problem by introducing \Q@AbstractSetOrBag@
and making both \Q@Bag@ and \Q@Set@ inherit from it.
This looks unnatural, since \Q@Set@ would extend it without adding anything,
and we would be surprised to find a use of the type \Q@AbstractSetOrBag@.
Worst, if we was to constantly apply this mentalty, we would introduce a very high number
of abstract classes that are not supposed to be used as types, and that will clutter the 
public interface of our classes and our code project as a whole.

>>>>

In mainstream OO languages like Java, C++ or C\# subclassing 
implies subtyping. For example, in Java a subclass definition, such as:

\begin{lstlisting}
class A extends B {}
\end{lstlisting}

\noindent does two things at the same time:
1) {\bf inheriting} code from class \lstinline{B}; and {\bf creating 
a subtype} of \emph{B}. Therefore in a language like Java 
a subclass is \emph{always} a subtype of the extended class.

Historically, there has been a lot of focus on the importance of
separing subtyping from subclassing~\cite{cook}.  This is claimed to be
good for code reuse, design and reasoning. 
While there is no problem
in subtyping without subclassing, in most OO languages code reuse
(inheritance/extends) implies subtyping in a fundamental way because 
of what we call the \emph{self-reference problem}. The self-reference
problem is illustrated by the following (Java) code:

@@

%%Class \lstinline{B} does two things at the same time: 
%%1) it {\bf inherits} the method \lstinline{ma} from
%%\lstinline{A}; and 2) it creates a {\bf subtype} of \lstinline{A}.
We can see an invocation of \lstinline{ma} inside
\lstinline{mb}, where \lstinline{this} is of type \lstinline{B}. 
The execution will eventually call \lstinline{Util.m} with an
instance of \lstinline{B}. This can be correct only if \lstinline{B} is a subtype of
\lstinline{A}.

If Java were to support a mechanism to allow reuse/inheritance 
without introducing subtyping, such as:

\begin{lstlisting}[language=Java]
  class B inherits A{ int mb(){return this.ma();} }
\end{lstlisting}

\noindent Then an invocation of 
\lstinline{mb} would be type-unsafe (i.e. it would 
result in a run-time type error). 
Note that here the intention of using the imaginary keyword {\bf
  inherits} is to allow the code from \lstinline{A} to be inherited 
without \lstinline{B} becoming a subtype of \lstinline{A}. 
However this breaks type-safety. The problem is that the
self-reference \lstinline{this} in class \lstinline{B} has 
type \lstinline{B}. Thus, when \lstinline{this} is passed as an argument to 
the method \lstinline{Utils.m} as a result of the invocation of
\lstinline{mb}, it will have a type that is incompatible with the
expected argument of type \lstinline{A}.

What the self-reference problem illustrates is that adopting a more flexible
OO model where subclassing does not imply subtyping is not trivial. 
If we are to separate subclassing from subtyping then a more substantial
change in the language design is necessary.  Every OO language with
the minimal features exposed in the example (using \lstinline{this},
extends and method calls) is forced to accept that subclassing implies
subtyping.\footnote{ Note how this is true also in C++, where is
  possible to "extends privately". Such is a limitation of the
  visibility of subtyping but not over subtyping itself, and the
  former example would be accepted by C++ even if \lstinline{B} was to
  "privately extends" \lstinline{A}.}

\marco{Cite some work of bruceTHisType and show how he also fails.}

\begin{comment}
An iconic example on why connecting inheritance/code reuse and
subtpying is problematic is provided by the
historic\cite{LaLonde:1991:SSS:110673.110679}: A reasonable
implementation for a \Q@Set@ may be easy to extend into a \Q@Bag@
keeping tracks of how many times an element occurs.  We would just add
some state and override a couple of methods.\bruno{Are we going to
  present this example in the paper solved in 42, for example? 
I think I would expect to see it.}

However, our subclassing would break Liskov substitution principle (LSP)~\cite{martin2000design}: not all bags are sets!
Of course, one could retroactivelly fix this problem by introducing \Q@AbstractSetOrBag@
and making both \Q@Bag@ and \Q@Set@ inherit from it.
This looks unnatural, since \Q@Set@ would extend it without adding anything,
and we would be surprised to find a use of the type \Q@AbstractSetOrBag@.
Worst, if we was to constantly apply this mentalty, we would introduce a very high number
of abstract classes that are not supposed to be used as types, and that will clutter the 
public interface of our classes and our code project as a whole.
\end{comment}
